A subjective Bayesian framework for synthesizing deep uncertainties in climate risk management

### 2022-11-23 papers/earths-future/submission_03.tex
reviewer response

In this paper we offer a conceptual step towards bridging this divide by presenting a framework that is designed to combine the strengths of both approaches.
In the first step, exploratory or bottom-up modeling is used to build insight and identify potential system vulnerabilities \cite{moallemi_exploratory:2020,bankes:1993,brown_decisionscaling:2012}.
In the second step, we synthesize the results of exploratory modeling to formally estimate performance metrics and trade-offs using probabilistic models for uncertainty.
Drawing from the literature on building predictive models when all models are wrong \cite{box_sciencestatistics:1976,gelman_philosophy:2013,Piironen:2017eh}, we interpret these probability distributions not as statements of fact, but rather as a self-consistent framework for reasoning about how different assumptions lead to different inferences.
An advantage of our approach is that it facilitates computationally efficient analysis of how alternative probabilistic models would affect estimated performance metrics and trade-offs.

@@

In this section we introduce a conceptual framework and notation for decision analysis under deep uncertainty.
We then apply this framework to understand exploratory modeling (\cref{sec:analysis-explore}) and scenario-conditional probabilistic analysis (\cref{sec:analysis-condition}).
In \cref{sec:analysis-synthesize} we discuss why synthesizing across multiple scenarios is necessary.
In \cref{sec:reweighting} we provide a formal method for synthesizing across scenarios.

It is common in climate risk management to work with an ensemble of trajectories of climate, economic, and other key variables, which we refer to as \glspl{sow}.
Often these \glspl{sow} are computationally expensive to generate, and thus it is important to use them efficiently.
Following \cref{fig:flowchart}, we consider using $J$ \glspl{sow}, $\vb{s} = \qty{s_1, s_2, \ldots, s_J}$, to evaluate $I$ candidate decisions, $\vb{x} = \qty{x_1, x_2, \ldots, x_I}$.
For each \gls{sow} $s_j \in \vb{s}$ and decision $x_i \in \vb{x}$ we use a system model $f$ (comprised of multiple components) to calculate a set of metrics describing the performance of decision $x_i$ on \gls{sow} $s_j$, which we denote $u_{ij} = f(x_i, s_j)$.
While we assume for simplicity that the decision space is known and finite, this approach could be coupled to a policy search model that proposes candidate decisions.

@@

One way to interpret an ensemble of \glspl{sow} is as {iid} draws from some probabilistic data generating process.
This commonly arises when a single deep uncertainty (\eg, an emissions pathway) is used as an input for a stochastic model.
We illustrate this concept in boxes (d) and (e) of \cref{fig:flowchart}, denoting the particular scenario $M_k$.
By treating the \glspl{sow} as \gls{iid} draws from $M_k$, the set of outcomes $u_{i, j}$ can be interpreted as \gls{iid} draws from the conditional distribution over outcomes, $p(u | x_i, M_k)$.
This ``scenario-conditional'' probabilistic interpretation of \glspl{sow} allows for fully probabilistic quantification of uncertainty and optimization, conditional on a particular scenario.
For example, \citeA{fletcher_learning:2019} uses stochastic dynamic programming to quantify the value of flexibility in water resources planning.
However, only a single \gls{rcp} scenario (\gls{rcp} 8.5) is used.
While the analysis could be repeated for other \gls{rcp} scenarios, the scenario-conditional analysis framework only qualitatively characterize uncertainty \emph{between} scenarios \cite{wong_nola:2017,ruckert_coastal:2019,sharma_rcp:2021}.

We can also characterize \gls{dmdu} methods that sample a set of parameters from fixed ranges as scenario-conditional analyses.
For example, \citeA{sriver_sealevel:2018} sample parameters describing the rate of \gls{slr} across a range of values to inform coastal adaptation.
Similarly, \citeA{trindade_waterpathways:2020} checks the performance of candidate decisions against an ensemble of synthetic time series of streamflow, water demand, and other parameters by sampling parameters that transform the available data over a plausible range.
Analyses that use this methodology are implicitly assuming a single probabilistic model in which different variables are drawn from independent Uniform distributions.
Many limitations of Uniform and other noninformative priors have been documented in the literature, including (i) that they can induce unrealistic implicit priors over functions of parameters and (ii) results are sensitive to the parameterization of a given process \cite{seaman_noninformative:2012}.
Yet while replacing Uniform distributions with alternatives such as maximum-entropy distributions can address some of these challenges \cite<\eg,>{gupta_prior:2022}, subjective modeling choices remain necessary.
Our primary concern here is not that these subjective modeling assumptions are wrong -- this is, almost surely, inevitable -- but that when these assumptions are opaque and presented without critique or validation \cite<see>[regarding the importance of iterative critique]{gelman_workflow:2020} they may lead to inscrutable decision processes and poor decisions.

@@

Scenario-conditional probabilistic analysis allows for uncertainty quantification and optimization, and is valuable in many copntexts.
However, scenarios are often explicitly provided without probabilities or likelihoods \cite<\eg, the \acrlongpl{ssp}>{vanvuuren_conditional:2008}.
Thus, any such analysis is silent on the question of how to combine information across different scenarios.
We term this the ``multiple \gls{pdf} problem''.
Decision making around the multiple \gls{pdf} problem is susceptible to the cognitive biases that interfere with decision-making under uncertainty more generally \cite{tversky_judgment:1974,morgan_uncertainty:1990,srikrishnan_probabilistic:2022}.
For example, while many analyses treat all scenarios as equally likely, this is often inconsistent with available information and can lead to poor decisions and outcomes \cite{wigley_equallylikely:2001,ho_scenarios:2019,hausfather_scenarios:2020}.
Other analyses suggest using the worst-case scenario as a conservative measure.
However, this approach is also problematic, since (i) there are no fundamental limits on what constitutes a worst-case scenario and (ii) improving performance under unlikely worst-case scenarios may lead to substantially impaired performance under more likely scenarios, which may or may not be acceptable to relevant stakeholders.
There is thus a critical need for synthesizing insights across multiple probabilistic scenarios.

@@

In this section we provide a formal method for synthesizing across multiple scenarios to inform climate risk management.
Our objective is to develop a framework that (i) is conceptually and practically amenable to exploratory modeling; (ii) makes subjective modeling choices explicit and transparent; and (iii) allows decision analysts to estimate a probability distribution over outcomes.

@@

Doing so requires a probabilistic model for this low-dimensional representation of the \glspl{sow},  $p(\psi)$.
We denote this model $p_\mathrm{belief}$ to emphasize that it represents a subjective belief about the \glspl{sow}, rather than an objectively verifiable choice.
In general we do not expect that stakeholders and experts will agree on $p_\mathrm{belief}$ because there is not, even conceptually, an objectively correct choice \cite{oreskes_verification:1994,walker_deep:2013}.
However, we posit that since we cannot be ``right,'' it is valuable to maximize the transparency of our implicit probabilistic assumptions, and suggest that writing down an explicit model for $p_\mathrm{belief}$ supports this aim.
Choices for $p_\mathrm{belief}$ can be drawn from many sources, including expert elicitation or  results of previous analyses.
These models can be interpreted as prior beliefs about \gls{slr} that could be incorporated into a Bayesian analysis as additional data is collected in the future, and thus can draw from literature on Bayesian prior selection and prior predictive checks \cite{gelman_workflow:2020}.

@@

\subsection{Prior over \glsentrylong{slr}}\label{sec:case-priors}

@@

We use a Gamma distribution for all three priors, parameterized following eq.~(S4).
The distributions were chosen to be illustrative, rather than to reflect any particular scientific consensus.
The Gamma distribution is a flexible distribution that can be used to model skewed, lower-bounded distributions, making it an appropriate choice for modeling subjective uncertainty about \gls{slr}.
\Cref{tab:slr-priors} specifies the parameters of these distributions, as well as some quantiles of the distributions.
Their \glspl{pdf} are also plotted in \cref{fig:lsl-priors-weights}(A).

@@

We developed these priors for didactic purposes, to illustrate a range of possible beliefs.
We can compare them, for example, with analysis published by \gls{noaa}, which project \SIlist{1.94;2.62;4.27;5.25;6.89}{ft} for the low, intermediate, low intermediate, intermediate high, and high scenarios, respectively \cite[table.~2.4]{sweet_slr:2022}.
We can also compare to the analyses of \citeA{sriver_sealevel:2018} which uses a rescaled Beta distribution with bounds of \SIrange{0.83}{8.2}{ft} and a most plausible estimate of \SI{3.1}{ft}.
Our samples bound all of these estimates.

>>>>

In this paper we offer a conceptual step towards bridging this divide by presenting a framework that is designed to combine the strengths of both approaches.
In the first step, exploratory or bottom-up modeling is used to build insight and identify potential system vulnerabilities \cite{moallemi_exploratory:2020,bankes:1993,brown_decisionscaling:2012}.
In the second step, we integrate exploratory ensembles of deep uncertainties into a single probabilistic representation (we refer to this as ``synthesizing'' deep uncertainties) to formally estimate performance metrics and trade-offs using subjective probability distributions.
Drawing from the literature on building predictive models when all models are wrong \cite{box_sciencestatistics:1976,gelman_philosophy:2013,Piironen:2017eh}, we interpret these probability distributions not as statements of fact, but rather as a self-consistent framework for reasoning about how different assumptions lead to different inferences.
An advantage of our approach is that it facilitates computationally efficient analysis of how alternative probabilistic models would affect estimated performance metrics and trade-offs.

@@

In this section we introduce a conceptual framework and notation for decision analysis under deep uncertainty.
Many bottom-up exploratory modeling frameworks used in climate risk management and related fields use a system model ($f$ in \cref{fig:flowchart})  to characterize the system's response to a wide range of plausible futures, often called \glspl{sow} ($\vb{s}$ in \cref{fig:flowchart}).
This analysis is often used to explore vulnerabilities and build knowledge about the system \cite{bankes:1993}, and in general exploratory modeling frameworks aim to avoid making explicit judgments about the relative likelihood of different futures.

However, as discussed in \cref{sec:introduction}, estimates of trade-offs between desired performance metrics (\eg, cost and reliability) depend on probabilistic models of uncertainty.
In this paper we present a method for integrating the \glspl{sow} used in exploratory modeling into a formal decision analytic framework using a subjective probability distribution over the space of possible futures, which is used to infer implicit weights over the \glspl{sow}.
This approach is particularly suited for problems where the \glspl{sow} are generated from or conditioned on specific scenarios (\eg, \gls{rcp} scenarios) or where there are multiple models of the underlying processes (\eg, multiple models of the response of local sea levels to global temperature), which can lead to the ``multiple \gls{pdf} problem'' (discussed in \cref{sec:analysis-condition}).
A motivating advantage is that it makes assumptions about the likelihoods of different \glspl{sow} transparent to decision-makers.

@@

One way to interpret an ensemble of \glspl{sow} is as {iid} draws from some probabilistic data generating process.
This commonly arises when a single deep uncertainty (\eg, an emissions pathway) is used as an input for a stochastic model.
To clarify language, we draw a distinction between a \gls{sow}, which is a single realization of a possible future, and a \emph{scenario}, which is a model that can be used to generate \glspl{sow}.
These terms are defined relative to the problem at hand; for example, a global emissions trajectory may be a \gls{sow} from an integrated assessment model, but when it is used as a deterministic input to an ice sheet model we would describe it as a scenario.

We illustrate this distinction in boxes (d) and (e) of \cref{fig:flowchart}, denoting the particular scenario $M_k$.
We assume that each scenario is probabilistic, that is that \glspl{sow} are drawn \gls{iid} from $M_k$, the set of outcomes $u_{i, j}$ can be interpreted as \gls{iid} draws from the conditional distribution over outcomes, $p(u | x_i, M_k)$.
This ``scenario-conditional'' probabilistic interpretation of \glspl{sow} allows for fully probabilistic quantification of uncertainty and optimization, conditional on a particular scenario.
For example, \citeA{fletcher_learning:2019} uses stochastic dynamic programming to quantify the value of flexibility in water resources planning.
However, only a single \gls{rcp} scenario (\gls{rcp} 8.5) is used.
While the analysis could be repeated for other \gls{rcp} scenarios, the scenario-conditional analysis framework only qualitatively characterize uncertainty \emph{between} scenarios \cite{wong_nola:2017,ruckert_coastal:2019,sharma_rcp:2021}.

We can also apply this theoretical lens to examine the approach, common in \gls{dmdu} applications, of sampling parameters from a set of fixed ranges.
The \emph{scenario} in this case is thus the choice of bounds on the parameters; it is consistent with our above definition of a scenario because it provides a probabilistic model that can be used to sample \glspl{sow}.
For example, \citeA{sriver_sealevel:2018} sample parameters describing the rate of \gls{slr} across a range of values to inform coastal adaptation.
Similarly, \citeA{trindade_waterpathways:2020} checks the performance of candidate decisions against an ensemble of synthetic time series of streamflow, water demand, and other parameters by sampling parameters that transform the available data over a plausible range.
Analyses that use this methodology are implicitly assuming a single probabilistic model in which different variables are drawn from independent Uniform distributions.
Many limitations of Uniform and other noninformative priors have been documented in the literature, including (i) that they can induce unrealistic implicit priors over functions of parameters and (ii) results are sensitive to the parameterization of a given process \cite{seaman_noninformative:2012}.
Yet while replacing Uniform distributions with alternatives such as maximum-entropy distributions can address some of these challenges \cite<\eg,>{gupta_prior:2022}, subjective modeling choices remain necessary.
Our primary concern here is not that these subjective modeling assumptions are wrong -- this is, almost surely, inevitable -- but that when these assumptions are opaque and presented without critique or validation \cite<see>[regarding the importance of iterative critique]{gelman_workflow:2020} they may lead to inscrutable decision processes and poor decisions.

@@

Scenario-conditional probabilistic analysis allows for uncertainty quantification and optimization, and is valuable in many contexts.
However, scenarios are often explicitly provided without probabilities or likelihoods \cite<\eg, the \acrlongpl{ssp}>{vanvuuren_conditional:2008}.
Thus, any such analysis is silent on the question of how to combine information across different scenarios.
We term this the ``multiple \gls{pdf} problem''.
Decision making around the multiple \gls{pdf} problem is susceptible to the cognitive biases that interfere with decision-making under uncertainty more generally \cite{tversky_judgment:1974,morgan_uncertainty:1990,srikrishnan_probabilistic:2022}.
For example, while many analyses treat all scenarios as equally likely, this is often inconsistent with available information and can lead to poor decisions and outcomes \cite{wigley_equallylikely:2001,ho_scenarios:2019,hausfather_scenarios:2020}.
Other analyses suggest using the worst-case scenario as a conservative measure.
However, this approach is also problematic, since (i) there are no fundamental limits on what constitutes a worst-case scenario and (ii) improving performance under unlikely worst-case scenarios may lead to substantially impaired performance under more likely scenarios, which may or may not be acceptable to relevant stakeholders.
There is thus a critical need for synthesizing insights across multiple probabilistic scenarios.

@@

In this section we provide a formal method for integrating exploratory ensembles of deep uncertainties into a single probabilistic representation.
Our objective is to develop a framework that (i) is conceptually and practically amenable to exploratory modeling; (ii) makes subjective modeling choices explicit and transparent; and (iii) allows decision analysts to estimate a probability distribution over outcomes.

@@

Implementing this approach requires choosing a probability distribution for this low-dimensional representation of the \glspl{sow},  $p(\psi)$, reflecting subjective belief about the \glspl{sow}.
We denote this $p_\mathrm{belief}(\psi)$ to emphasize that it represents a subjective belief about the \glspl{sow}, rather than an objectively verifiable choice.
In general we do not expect that stakeholders and experts will agree on $p_\mathrm{belief}$ because there is not, even conceptually, an objectively correct choice \cite{oreskes_verification:1994,walker_deep:2013}.
However, we posit that since we cannot be ``right,'' it is valuable to maximize the transparency of our implicit probabilistic assumptions, and suggest that writing down an explicit model for $p_\mathrm{belief}$ supports this aim.
Choices for $p_\mathrm{belief}$ can be drawn from many sources, including expert elicitation or  results of previous analyses.
These models can be interpreted as prior beliefs about \gls{slr} that could be incorporated into a Bayesian analysis as additional data is collected in the future, and thus can draw from literature on Bayesian prior selection and prior predictive checks \cite{gelman_workflow:2020}.

@@

\subsection{Subjective probability distributions for \glsentrylong{slr}}\label{sec:case-priors}

@@

We use a Gamma distribution for all three probability distributions, parameterized following eq.~(S4).
The distributions were chosen to be illustrative, rather than to reflect any particular scientific consensus.
The Gamma distribution is a flexible distribution that can be used to model skewed, lower-bounded distributions, making it an appropriate choice for modeling subjective uncertainty about \gls{slr}.
\Cref{tab:slr-priors} specifies the parameters of these distributions, as well as some quantiles of the distributions.
Their \glspl{pdf} are also plotted in \cref{fig:lsl-priors-weights}(A).

@@

We developed these subjective distributions for didactic purposes, to illustrate a range of possible beliefs.
We can compare them, for example, with analysis published by \gls{noaa}, which project \SIlist{1.94;2.62;4.27;5.25;6.89}{ft} for the low, intermediate, low intermediate, intermediate high, and high scenarios, respectively \cite[table.~2.4]{sweet_slr:2022}.
We can also compare to the analyses of \citeA{sriver_sealevel:2018} which uses a rescaled Beta distribution with bounds of \SIrange{0.83}{8.2}{ft} and a most plausible estimate of \SI{3.1}{ft}.
Our samples bound all of these estimates.
